Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2018-02-13 paperIEEE.tex
Preflight check
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@125 db7c848b-7755-4df2-b763-8796797afa52

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks, or work packages,
are queued and processed by all source channels by polling a server-central
queue. Prefetching ensures that the task communication overlaps with rendering.
As shown in~ \cite{SPEP:16}, these modes can provide better performance due to
being inherently load-balanced, as long as there is an insignificant overhead
for the render task setup. This mode is transparent to \textsf{Equalizer}
applications.

@@

@@

Due to the master-slave model of data distribution, multicast can be used to
optimize the transmission time of data. If the contributing nodes share a
multicast session, and more than one slave instance is mapped, \textsf{Collage}
automatically uses the multicast connection to send the new version
information.

@@

\textsf{Collage} implements a few generic distributed objects which are used by
\textsf{Equalizer}
and applicatons. A barrier is a distributed barrier primitive used for software
swap barriers in \textsf{Equalizer} (\sref{sec:swap}). Its implementation
follows a
simple master-slave approach, which has shown to by sufficient for this use
case. Queues are single producer, multiple consumer FIFOs. To hide network
latencies, the consumers use prefetching of queue items into a local queue. They
are used for tile and chunk compounds (\sref{sec:tile}).

@@

This section presents new experiments, complementing the results of previous
publications~\cite{EP:07, EMP:09, MEP:10, EEP:11, EBAHMP:12, HBBES:13, deflect,
  SPEP:16}. The first part summarizes rendering performance over all
decomposition modes with a few representative workloads. The second part
analyses data distribution performance, in particular how the various
optimizations in \textsf{Collage} affect complex operations in realistic
scenarios.

@@

We tested the decomposition modes with both volume data (\fig{fBenchmarkEVolve})
and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us
to continuously modify the rendering load the system has to cope with. In both cases
the scene comprised of two rows of instantiated, identical models with 30 models
in total. Rendering was performed at an output resolution of $2560\times1440$. The camera is
initially placed in the center of the scene, between the two rows
(\fig{fBenchmarkEVolve} and \fig{fBenchmarkEqPly}), rendering only half of the
model instances. It is then moved backward over the duration of 800 frames,
steadily increasing the rendering load by revealing more models, until all 30
instances are visible.

We investigated the scalability of individual decomposition modes by running the
same experiment using a varying number of render nodes (2-9), as well as one
dedicated application/display node. We subsequently summed up the duration of
all rendered frames for each run (\fig{fCompounds}), backing up along the alley of models.
We provide the results of the load balancer providing the better result for each application and compound.\FIXME{what does this last sentence mean?}
The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
load-grid-driven \textsf{load\_equalizer}, except for sort-first volume
rendering where the load in the region of interest is relatively uniform. This
counterintuitive result once more proves that simple algorithms often outperform
theoretically better, but more complex implementations.

\FIXME{oscillation, pixel, clearer}
\FIXME{I'm not sure I read Fig.10 the same way. Left it seems that tile\_equalizer is the best for polygonal rendering, it's constantly faster than the tree\_equalizer mentioned in the paragraph above.}

@@

\begin{figure}[ht]\center
  \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
  \caption{\label{fBenchmarkEVolve} Flower alley with 30 volume models of $1024^3$ unsigned bytes (1\,GB) each.}
\end{figure}

\begin{figure}[ht]\center
  \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56\,M triangles (988\,MB) each.
\end{figure}

@@

The data distribution benchmarks have been performed on a cluster using
dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory per node, 10\,GBit/s
Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release
mode settings to compile the software stack.

To benchmark the data distribution we used two datasets: The David statue at 2\,mm
resolution (as in \fig{fBenchmarkEqPly}, but in 2\,mm resolution) and 3D volumes
of spike frequencies of an electrical simulation of three million neurons
(\fig{fVolumes}).

The PLY file is converted into a k-d tree for fast view frustum culling and
rendering, and the resulting data structure is serialized in binary form for
data transmission.

\begin{figure}[ht]\center
  \includegraphics[height=.44\columnwidth]{images/spikes512}\hfil
  \includegraphics[height=.44\columnwidth]{images/beechnut}
  \caption{\label{fVolumes}Volumes used for object distribution benchmarks:
    spike frequencies of a three million neuron electrical simulation at
    512x437x240 unsigned byte (51\,MB, left) and an MicroCT scan of a beechnut at
    1024x1024x1546 unsigned short (3\,GB, right) resolution}
\end{figure}

The spike frequency volumes aggregate the number of spikes which happened within
a given voxel over a given time. The absolute spike count is renormalized to an
unsigned byte value during creation. Higher densities in the volume represent
higher spiking activity in the voxel.

@@

A critical piece for data distribution performance are the characteristics of
the data compression algorithm. Our microbenchmark compresses a set of binary
files, precalculating the speed and compression ratio of the various engines.
\fig{fCompressor} shows the compression and decompression speed in gigabyte per
second as well as the size of the compressed data relative to the uncompressed
data. The different ZSTD$x$ engines use the ZStandard compression library at
compression level $x$. The measurements were performed on a single, isolated
node.

@@

Finally, we compare data distribution speed using different protocols. In this
benchmark, data synchronization time of the Spike volume data set is measured,
as in \fig{fSpikesDist} (middle). Buffering is enabled, and compression is
disabled to focus on the raw network performance. \fig{fIFDist} shows the
performance using various protocols.

Unsurprisingly, TCP over the faster InfiniBand link outperforms the cheaper ten
gigabit ethernet link by more than a factor of two. Unexpectedly, the native
RDMA connection performs worse, even though it outperforms IPoIB in a simple
peer-to-peer connection benchmark. This needs further investigation, but we
suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
is not well suited for remote DMA semantics; that is, one needs to design the
network API around zero-copy semantics with managed memory for modern high-speed
transports. Both Infiniband connections show significant measurement jitter.

@@

\begin{figure*}[ht]\center
  \includegraphics[width=\textwidth]{images/compressorDetail}
  \caption{\label{fCompressorDetail}Data compression for PLY data (left, David
    statue 2\,mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
\end{figure*}

\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/daviddist}
  \caption{\label{fDavidDist}Object mapping (left) and data synchronization
    time (middle, detail view right) for the David 2\,mm data set}
\end{figure*}
\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/spikesdist}
  \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
    time (middle, detail view right) for the Spike data set in
    \fig{fVolumes}, left}
\end{figure*}
\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/beechnutdist}
  \caption{\label{fBeechnutDist}Object mapping (left) and data synchronization
    time (middle, detail view right) for the beechnut data set in
    \fig{fVolumes}, right}
\end{figure*}
\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/ifdist}
  \caption{\label{fIFDist}Object synchronization using different network
    transports}
\end{figure*}

>>>>

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks, or work packages,
are queued and processed by all source channels by polling a server-central
queue. Prefetching ensures that the task communication overlaps with rendering.
As shown in~\cite{SPEP:16} and the results, these modes can provide better
performance due to being implicitly, i.e., inherently load-balanced, as long as
there is an insignificant overhead for the render task setup. This mode is
transparent to \textsf{Equalizer} applications.

@@

\subsubsection{Dynamic Work Packages}

The package equalizers implement client-affinity models for tile and chunk
compounds (\sref{sec:tile}). A \textsf{tile\_equalizer} and
\textsf{chunk\_equalizer} create the packages and change the assignment of them
to individual nodes, based on an affinity model specified in the equalizer.
In~\cite{SPEP:16}, we explore this approach in detail.

@@

Due to the master-slave model of data distribution, multicast is used to
optimize the transmission time of data. If the contributing nodes share a
multicast session, and more than one slave instance is mapped, \textsf{Collage}
automatically uses the multicast connection to send the new version information.

@@

\textsf{Collage} implements a few generic distributed objects which are used by
\textsf{Equalizer} and other applicatons. A barrier is a distributed barrier
primitive used for software swap barriers in \textsf{Equalizer}
(\sref{sec:swap}). Its implementation follows a simple master-slave approach,
which has shown to by sufficient for this use case. Queues are single producer,
multiple consumer FIFOs. To hide network latencies, consumers prefetch queue
items into a local queue. They are used for tile and chunk compounds
(\sref{sec:tile}).

@@

This section presents new experiments, complementing the results of previous
publications~\cite{EP:07, EMP:09, MEP:10, EEP:11, EBAHMP:12, HBBES:13, deflect,
SPEP:16}. The first part summarizes rendering performance over all decomposition
modes with a few representative workloads. The second part analyses data
distribution performance, in particular how the optimizations in
\textsf{Collage} perform in realistic scenarios.

@@

We tested the decomposition modes with both polygonal data and volume data
(\fig{fBenchmarks} (middle and left)), using test scenes that allowed to adapt
the rendering load the system has to cope with. In both cases the scene is
comprised of two rows of instantiated, identical models with 30 models in total.
Rendering was performed at an output resolution of $2560\times1440$. The camera
is initially placed in the center of the scene, between the two rows, rendering
only half of the model instances. It is then moved backward over the duration of
800 frames, steadily increasing the rendering load by revealing more models,
until all 30 instances are visible.

We investigated the scalability of individual decomposition modes by running the
same experiment using a varying number of render nodes (2-9) and one dedicated
application/display node. We subsequently summed up the duration of all rendered
frames for each run (\fig{fCompounds}).

For sort-first and sort-last rendering we present static and load-balanced task
decomposition. For readability, we only present the results of the equalizer
(load or tree) providing the better performance for each application.
Unsurprisingly, static decompositions perform worse over load-balanced
compounds. Sort-first polygon rendering exhibits oscillations in performance as
nodes are added to the task, due to unfavorable assignment of a tile with a high
work load on odd node counts. Static sort-last volume rendering has a similar
oscillation behaviour \FIXME{due to @David?}.

The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
load-grid-driven \textsf{load\_equalizer}, except for sort-first volume
rendering where the load in the region of interest is relatively uniform. This
counterintuitive result seems to again confirm that simple algorithms often
outperform theoretically better, but more complex implementations. On the other
hand, the \textsf{tile\_equalizer} often outperforms \textsf{tree\_equalizer}.
This suggests that the underlying implicit load balancing can be superior to the
explicit methods of \textsf{load\_equalizer} and \textsf{tree\_equalizer} in
high load situations, where the additional overhead of tile generation and
distribution is more justified. The relative simple nature of our benchmark
application's rendering algorithms is also favoring work packages, since it has
a near-zero static overhead per rendering pass.

Last, but not least, we also provide scalability results for pixel compounds.
While naturally load-balanced, pixel compounds only scale fill rate and not
geometry processing. Consequently, pixel compounds provide better performance
for volume rendering, and a predictable scaling behaviour for both.

@@

\begin{figure*}[ht]\center
    \includegraphics[height=.17\textwidth]{images/benchmark_eVolve}
    \includegraphics[height=.17\textwidth]{images/benchmark_eqPly2}
    \includegraphics[height=.17\textwidth]{images/spikes512}
    \includegraphics[height=.17\textwidth]{images/beechnut}
  \caption{\label{fBenchmarks}Benchmark data: Flower alley with 30 volume models
  of $1024^3$ unsigned bytes (1\,GB) each (left); David alley with 30 statues of
  56\,M triangles (988\,MB) each (middle); Volumes used for object distribution
  benchmarks (right): spike frequencies of a three million neuron electrical
  simulation at 512x437x240 unsigned byte (51\,MB) and an MicroCT scan of a
  beechnut at 1024x1024x1546 unsigned short (3\,GB) resolution}
\end{figure*}

@@

The data distribution benchmarks have been performed on a cluster using dual
processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory per node,
10\,GBit/s Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake
3.2 release mode settings to compile the software stack. To benchmark the data
distribution we used two datasets: The David statue at 2\,mm resolution (as in
\fig{fBenchmarks (middle)}, but in 2\,mm resolution) and 3D volumes of spike
frequencies of an electrical simulation of three million neurons
(\fig{fBenchmarks} (right)).

The PLY file is converted into a k-d tree for fast view frustum culling and
rendering, and the resulting data structure is serialized in binary form for
data transmission. The spike frequency volumes aggregate the number of spikes
which happened within a given voxel over a given time. The absolute spike count
is renormalized to an unsigned byte value during creation. Higher densities in
the volume represent higher spiking activity in the voxel.

@@

A critical piece for data distribution performance are the characteristics of
the data compression algorithm. Our microbenchmark compresses a set of binary
files, precalculating the speed and compression ratio of the various engines.
\fig{fCompressor} shows the compression and decompression speed in gigabyte per
second as well as the size of the compressed data relative to the uncompressed
data. The ZSTD$x$ engines use the ZStandard compression library at compression
level $x$. The measurements were performed on a single, isolated node.

@@

Finally, we compare data distribution speed using different protocols. In this
benchmark, data synchronization time of the Spike volume data set is measured,
as in \fig{fSpikesDist} (middle). Buffering is enabled, and compression is
disabled to focus on the raw network performance. \fig{fIFDist} shows the
performance using various protocols. TCP over the faster InfiniBand link
outperforms the cheaper ten gigabit ethernet link by more than a factor of two.
Unexpectedly, the native RDMA connection performs worse, even though it
outperforms IPoIB in a simple peer-to-peer connection benchmark. This needs
further investigation, but we suspect the abstraction of a byte stream
connection chosen by \textsf{Collage} is not well suited for remote DMA
semantics; that is, one needs to design the network API around zero-copy
semantics with managed memory for modern high-speed transports. Both Infiniband
connections show significant measurement jitter.

@@

\begin{figure*}[ht]\center
  \includegraphics[width=\textwidth]{images/compressorDetail}
  \caption{\label{fCompressorDetail}Data compression for PLY data (left, David
    statue 2\,mm) and raw volumes shown in \fig{fBenchmarks} (right)}
\end{figure*}

\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/daviddist}
  \caption{\label{fDavidDist}Object mapping (left) and data synchronization
    time (middle, detail view right) for the David 2\,mm data set}
\end{figure*}
\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/spikesdist}
  \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
    time (middle, detail view right) for the Spike data set in
    \fig{fBenchmarks} (right)}
\end{figure*}
\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/beechnutdist}
  \caption{\label{fBeechnutDist}Object mapping (left) and data synchronization
    time (middle, detail view right) for the beechnut data set in
    \fig{fBenchmarks} (right)}
\end{figure*}
\begin{figure*}[p!]\center
  \includegraphics[height=.19\textheight]{images/ifdist}
  \caption{\label{fIFDist}Object synchronization using different network
    transports}
\end{figure*}
